Sequence Divergence as the main modulator of structure-sequence correlations

### 2014-11-03 manuscript/cordiv.tex
new data, analysis and figures added for proportion of variance explained.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   document style macros
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def\gtrsim{\mathrel{\hbox{\rlap{\hbox{\lower4pt\hbox{$\sim$}}}\hbox{$>$}}}}
\def\lessim{\mathrel{\hbox{\rlap{\hbox{\lower4pt\hbox{$\sim$}}}\hbox{$<$}}}}
\newcommand{\liso}{L_{iso}}
\newcommand{\eiso}{E_{iso}}
\newcommand{\epkz}{E_{p,z}}
\newcommand{\durz}{T_{90,z}}
\newcommand{\pbol}{P_{bol}}
\newcommand{\sbol}{S_{bol}}
\newcommand{\epk}{E_{p}}
\newcommand{\dur}{T_{90}}
\newcommand{\pph}{P_{50-300}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def\Title#1{\begin{center} {\Large #1 } \end{center}}
\def\Author#1{\begin{center}{ \sc #1} \end{center}}
\def\Address#1{\begin{center}{ \it #1} \end{center}}
\def\andauth{\begin{center}{and} \end{center}}
\def\submit#1{\begin{center}Submitted to {\sl #1} \end{center}}
\newcommand\pubblock{\rightline{\begin{tabular}{l} \pubnumber\\
         \pubdate  \end{tabular}}}
\newenvironment{Abstract}{\begin{quotation}  }{\end{quotation}}
%\newenvironment{Presented}{\begin{quotation} \begin{center}
%             PRESENTED AT\end{center}\bigskip
%      \begin{center}\begin{large}}{\end{large}\end{center} \end{quotation}}
\def\Acknowledgements{\bigskip  \bigskip \begin{center} \begin{large}
             \bf ACKNOWLEDGEMENTS \end{large}\end{center}}

@@

    Other structural properties have also been identified recently and proposed to influence or explain the site-specific evolutionary variations of proteins. Among the simplest properties is a measure of local density of the protein defined as the number of amino acids within a spherical neighborhood of a specific residue of interest, named residue {\it contact number (CN)}(xx). Variants of this quantity that eliminate the free-parameter (i.e., radius) in the definition of CN have also been proposed (xx) and have been shown to correlate better with sequence evolutionary rates (xx). In a recent work Echave et al. xx presented a biophysical model that links the thermodynamic stability changes due to mutations at sites in proteins ($\Delta\Delta G$) to the rate at which mutations accumulate at the corresponding sites over evolutionary time. They find that the variations in the free energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates, comparable to the predictive powers of other structural variables such as residue solvent accessibility and contact number.
    \\

    Not all proteins exhibit the same correlation strength and association between sequence variation and structural properties. By analyzing a data set of $216$ monomeric enzymes, Yeh et al. xx find a wide range of $\rho\sim0.1-0.8$ for the distributions of the correlations of sequence variability with two site-specific properties: contact number and residue solvent accessibility. Similarly, Echave et al. xx find a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by $\delta\delta G$ with evolutionary rates. Interestingly however, it appears that sequence-structure correlations also tend to correlate strongly with each other, as illustrated in Figure \ref{fig:ssscor}. This implies that for a given protein, the correlation strength of a specific structural property with evolutionary rates can serve as a proxy for the correlation strength of other structural properties with sequence evolutionary rates.
    \\

@@

    The results presented in this work are based on data set of 213 monomeric enzymes randomly picked from the Catalytic Site Atlas 2.2.11 (Porter et al. 2004) with protein sizes in the sample ranging from $95$ to $1287$ and including representatives of all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al. 1995). To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected for each protein from the {\it Clean Uniprot} database following the ConSurf protocol (Goldenberg et al. 2009; Ashkenazy et al. 2010). Sequence alignments were then constructed using amino-acid sequences with MAFFT (Katoh et al. 2002, 2005), specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data. The alignments were then used to calculate the site-specific evolutionary rates for each individual protein in dataset. To do so, we relied on two independent methods of measuring sequence variability measure. First, we calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:

@@

    All data including a list of 213 proteins and their properties together with Python, R and Fortran codes written for data reduction and analysis are publicly available to view and download at \url{https://github.com/shahmoradi/cordiv}.

@@

        As explained in the previous section, there is a high level of redundancy in the initial set of collected protein properties. In particular, depending on the set of atomic coordinates used, there are $7$ different measures for some residue characteristics such as the residue Weighted Contact Number, Bfactor and Voronoi cell properties. This in turn results in a large set of secondary variables at pdb-level that basically measure the same protein characteristics, but with different strengths.  Therefore, in order to eliminate redundant variables from dataset, we first compare the predictive power of different measures of residue characteristics based on the set of atomic coordinates used.
        \\

        For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the best correlation with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates. Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, $\Delta\Delta$ entropy, sequence entropy and evolutionary rates ($r4sJC$) by a median Spearman correlation difference of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).
        \\

        For the measure of local flexibility in proteins (Bfactor) we similarly find that among all $7$ representative measures of site Bfactors, the average of Bfactor values over all heavy atoms of each individual side chain (bfSC) results in the best correlations with other structural and sequence properties. Specifically, bfSC outperforms the commonly used $C_\alpha$ Bfactor (bfCA) in predicting RSA, $\Delta\Delta$ entropy, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
        \\

@@

        Based on the observations described in the previous paragraphs, we keep only variables measured based on average side chain properties and coordinates throughout the rest of the analysis and omit all other possible measures that show only weaker correlations with other site-specific characteristics.
        \\

    \subsection*{Identifying Determinants of Sequence-Structure Correlation Strengths}

        In order to identify the potential contributing factors to the strength of sequence--structure correlations, we first employ one of the simplest nonparametric yet powerful tests of statistical dependence between variables in a collection of observations, that is, we construct the Spearman correlation matrix of all pdb-level structure and sequence properties. The choice of Spearman vs. the popular Pearson's correlation measure is to minimize the effects of any potential nonlinear relationships between the set of variables on the measures of associations between the variables.

        The exclusion of alternative measures of structural quantities as described in the previous section, results in a significant reduction in the number of pdb-level variables to be further analysed, from $$ to $$, without compromising generality and comprehensiveness of the analysis.

        Now with the reduced data set at hand, we attempt to identify the main determinants of each sequence-structure relation of interest using regularized regression methods (xx) on the original dataset of pdb-level properties. , also on its rank transformation to ensure non-linearities

@@

        -- most importantly WCN, RSA $\Delta\Delta G$ Entropy, and Voronoi cell volume and surface area --

@@

        \begin{figure}[tbh]
            \begin{center}
            \includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf} % Fig. 1
            \end{center}
            \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs. correlations of other structural properties with evolutionary rates. Detailed description of the structural properties is given in Section \label{sec:mam}. The red lines in each plot represent equality line. It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures. For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
            \label{fig:seqent_structure_cors}
        \end{figure}

        \begin{figure}[tbh]
            \begin{center}
            \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_all_in_one.pdf} % Fig. 1
            \end{center}
            \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset. In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other. The capital letters in the variable names on the horizontal axis denote the set of atomic coordinates used to calculate WCN. The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
            \label{fig:best_wcn}
        \end{figure}

        \begin{figure}[tbh]
            \begin{center}
            \includegraphics[width=6.9in]{/best_bf/selected_variables/boxplot_bf_all_in_one.pdf} % Fig. 1
            \end{center}
            \caption{A comparison of the predictive power of different measures of Bfactor (BF) with different structure or sequence properties for 209 proteins in dataset. In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of atomic Bfactors on the horizontal axis are compared against each other. The capital letters in the variable names on the horizontal axis denote the representative atomic Bfactors. The variable $bfSC$ denotes the average coordinate over all heavy atoms in the side chain for each site in protein and the variable $bfAA$ denotes average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
            \label{fig:best_bf}
        \end{figure}

        \begin{figure}[tbh]
            \begin{center}
            \includegraphics[width=6.9in]{/best_vorvol/selected_variables/boxplot_vorvol_all_in_one.pdf} % Fig. 1
            \end{center}
            \caption{Should be likely modified to volume-area-length-sphericity measures instead.}
            \label{fig:best_voronoi}
        \end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   document style macros
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def\gtrsim{\mathrel{\hbox{\rlap{\hbox{\lower4pt\hbox{$\sim$}}}\hbox{$>$}}}}
\def\lessim{\mathrel{\hbox{\rlap{\hbox{\lower4pt\hbox{$\sim$}}}\hbox{$<$}}}}
\newcommand{\ddg}{$\Delta\Delta G~$}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def\Title#1{\begin{center} {\Large #1 } \end{center}}
\def\Author#1{\begin{center}{ \sc #1} \end{center}}
\def\Address#1{\begin{center}{ \it #1} \end{center}}
\def\andauth{\begin{center}{and} \end{center}}
\def\submit#1{\begin{center}Submitted to {\sl #1} \end{center}}
\newcommand\pubblock{\rightline{\begin{tabular}{l} \pubnumber\\
         \pubdate  \end{tabular}}}
\newenvironment{Abstract}{\begin{quotation}  }{\end{quotation}}
%\newenvironment{Presented}{\begin{quotation} \begin{center}
%             PRESENTED AT\end{center}\bigskip
%      \begin{center}\begin{large}}{\end{large}\end{center} \end{quotation}}
\def\Acknowledgements{\bigskip  \bigskip \begin{center} \begin{large}
             \bf ACKNOWLEDGEMENTS \end{large}\end{center}}

@@

    Other structural properties have also been identified recently and proposed to influence or explain the site-specific evolutionary variations of proteins. Among the simplest properties is a measure of local density of the protein defined as the number of amino acids within a spherical neighborhood of a specific residue of interest, named residue {\it contact number (CN)}(xx). Variants of this quantity that eliminate the free-parameter (i.e., radius) in the definition of CN have also been proposed (xx) and have been shown to correlate better with sequence evolutionary rates (xx). In a recent work Echave et al. xx presented a biophysical model that links the thermodynamic stability changes due to mutations at sites in proteins (\ddg) to the rate at which mutations accumulate at the corresponding sites over evolutionary time. They find that the variations in the free energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates, comparable to the predictive powers of other structural variables such as residue solvent accessibility and contact number.
    \\

    Not all proteins exhibit the same correlation strength and association between sequence variation and structural properties. By analyzing a data set of $216$ monomeric enzymes, Yeh et al. xx find a wide range of $\rho\sim0.1-0.8$ for the distributions of the correlations of sequence variability with two site-specific properties: contact number and residue solvent accessibility. Similarly, Echave et al. xx find a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg with evolutionary rates. Interestingly however, it appears that sequence-structure correlations also tend to correlate strongly with each other, as illustrated in Figure \ref{fig:ssscor}. This implies that for a given protein, the correlation strength of a specific structural property with evolutionary rates can serve as a proxy for the correlation strength of other structural properties with sequence evolutionary rates.
    \\

@@

    The results presented in this work are based on data set of 209 monomeric enzymes randomly picked from the Catalytic Site Atlas 2.2.11 (Porter et al. 2004) with protein sizes in the sample ranging from $95$ to $1287$ and including representatives of all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al. 1995). To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected for each protein from the {\it Clean Uniprot} database following the ConSurf protocol (Goldenberg et al. 2009; Ashkenazy et al. 2010). Sequence alignments were then constructed using amino-acid sequences with MAFFT (Katoh et al. 2002, 2005), specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data. The alignments were then used to calculate the site-specific evolutionary rates for each individual protein in dataset. To do so, we relied on two independent methods of measuring sequence variability measure. First, we calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:

@@

    All data including a list of 209 proteins and their properties together with Python, R and Fortran codes written for data reduction and analysis are publicly available to view and download at \url{https://github.com/shahmoradi/cordiv}.

@@

        As explained in previous section, there is a high level of redundancy in the initial set of collected protein properties. In particular, depending on the set of atomic coordinates used, there are $7$ different measures for some residue characteristics such as the residue Weighted Contact Number, Bfactor and Voronoi cell properties. This in turn results in a large set of secondary variables at pdb-level that basically measure the same protein characteristics, but with different strengths.  Therefore, in order to eliminate redundant variables from dataset, we first compare the predictive power of different measures of residue characteristics based on the set of atomic coordinates used.
        \\

        For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the best correlation with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates. Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates (r4sJC) by a median Spearman correlation difference of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).
        \\

        For the measure of local flexibility in proteins (Bfactor) we similarly find that among all $7$ representative measures of site Bfactors, the average of Bfactor values over all heavy atoms of each individual side chain (bfSC) results in the best correlations with other structural and sequence properties. Specifically, bfSC outperforms the commonly used $C_\alpha$ Bfactor (bfCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
        \\

@@

        Based on the observations described in the previous paragraphs, we keep only variables measured from average side chain properties and coordinates throughout the rest of the analysis and omit all other similar measures that show only weaker correlations with other site-specific characteristics. The exclusion of these alternative measures results in a significant reduction in the number of pdb-level variables to be further analysed, without compromising generality and comprehensiveness of the analysis.
        \\

    \subsection*{Sequence divergence as the main Determinant of Sequence-Structure Relation}

        In order to identify the potential contributing factors to the strength of sequence--structure correlations, we first employ one of the simplest nonparametric yet powerful tests of statistical dependence, that is, we construct the Spearman correlation matrix of all pdb-level structure and sequence properties. The choice of Spearman versus the popular Pearson's correlation measure is to minimize the effects of any nonlinear relationships between the variables on the strengths of the correlations. As summary of the most influential factors to structure-sequence relations is illustrated in Figure \ref{fig:}.

        Now with the reduced data set at hand, we attempt to identify the main determinants of each sequence-structure relation of interest, using regularized regression methods (xx) on the original dataset of pdb-level properties. , also on its rank transformation to ensure non-linearities

@@

        -- most importantly WCN, RSA \ddg Entropy, and Voronoi cell volume and surface area --

@@

        \begin{figure}[tbh]
            \begin{center}
            \includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf}
            \end{center}
            \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs. correlations of other structural properties with evolutionary rates. Detailed description of the structural properties is given in Section \label{sec:mam}. The red lines in each plot represent equality line. It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures. For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
            \label{fig:seqent_structure_cors}
        \end{figure}

        \begin{figure}[tbh]
            \begin{center}
            \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_all_in_one.pdf}
            \end{center}
            \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset. In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other. The capital letters in the variable names on the horizontal axis denote the set of atomic coordinates used to calculate WCN. The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
            \label{fig:best_wcn}
        \end{figure}

        \begin{figure}[tbh]
            \begin{center}
            \includegraphics[width=6.9in]{/best_bf/selected_variables/boxplot_bf_all_in_one.pdf}
            \end{center}
            \caption{A comparison of the predictive power of different measures of Bfactor (BF) with different structure or sequence properties for 209 proteins in dataset. In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of atomic Bfactors on the horizontal axis are compared against each other. The capital letters in the variable names on the horizontal axis denote the representative atomic Bfactors. The variable $bfSC$ denotes the average coordinate over all heavy atoms in the side chain for each site in protein and the variable $bfAA$ denotes average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
            \label{fig:best_bf}
        \end{figure}

        \begin{figure}[tbh]
            \begin{center}
            \includegraphics[width=6.9in]{/best_vorvol/selected_variables/boxplot_vorvol_all_in_one.pdf}
            \end{center}
            \caption{Should be likely modified to volume-area-length-sphericity measures instead.}
            \label{fig:best_voronoi}
        \end{figure}

        \begin{figure*}[tbh]
            \centering
            \begin{tabular}{cc}
            \includegraphics[width=3.4in]{/modulators_similarity_Sr4sJC_scors.pdf} & \includegraphics[width=3.4in]{/modulators_similarity_Sseqent_scors.pdf} \\
            \end{tabular}
            \caption{Proportion of explained variance of structure-sequence correlations by 182 pdb-level properties considered individually.}
            \label{fig:modsim}
        \end{figure*}
